Modeling the competing effects of the immune system and EMT on epithelial cancers

### 2019-09-16 Paper/CleanPaper.tex
Clean paper and premalignant/malignant

\section*{Abstract}
\sout{Preceding and d}\tcb{D}uring cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor \sout{incidence and} prognosis.
Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival.
Delaying \sout{tumorigenesis} \tcb{progression} depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
These results offer novel means to delay disease \sout{onset} \tcb{progression} by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.

@@

\tcr{In addition, the immune system plays important regulatory roles in pre-cancerous tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), \sout{particularly} \tcb{including} in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.}

@@

\tcr{Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.}

@@

While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor \tcr{initiation} \tcb{\cite{puisieux2014oncogenic}} and progression \cite{nieto2016emt}\tcb{\cite{peinado2007snail}}.

@@

\tcb{There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.} Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked. It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects} \sout{(which, under certain conditions can be called quiescence, though we avoid this term as it can be controversial, and is not a necessary assumption to the model developed below)}.

@@

We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that cancer will \sout{occur} \tcb{progress} and, if so, when.

@@

\tcb{

@@

}

@@

Global sensitivity analysis identifies parameters that are crucial for \sout{tumorigenesis} \tcb{progression}.
We then study these in more depth, focusing on the competing effects of EMT and of the immune system on \sout{tumorigenesis} \tcb{progression}: we find that EMT intricately regulates \sout{tumorigenesis} \tcb{progression} and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
\sout{To test these predictions, we analyze data from the Cancer Genome Atlas (TCGA), and find evidence for the synergistic effects predicted by the model for patients with pancreatic cancer.}

@@

\begin{table}[H]
\begin{tabular}{|p{\textwidth}|}
\hline
%%%
\tcb{
\begin{itemize}
\item TGF-$\beta$ is well-mixed within the TME
\item All cells are on the same ``clock'', i.e. they all decide at the same time what to do next
\item During a given cell cycle, exactly one event can happen to any one cell: proliferation, apoptosis, immune clearance, or rest
\item All tissue cells (mutated and non-mutated) are equally likely to encounter immune cells
\item Other spatial effects are ignored by the model, e.g. angiogenesis, hypoxia, invasive front, and chemotaxis
\item The initial cells are mutated, but are not able to form an aggressive cancer without further mutations
\end{itemize}
}
\\
\hline
\end{tabular}
\caption{\tcb{Basic assumptions of the model}}
\label{table:model_assumptions}
\end{table}

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
The agents in the model are the tissue cells \sout{with tumorigenic potential} \tcb{that have already formed a non-aggressive tumor yet have not progressed}.
\sout{These can either be mutation-free or, resulting from DNA damage during the cell cycle, can have any combination of three possible pathway mutations (shown in Fig. \ref{fig:ModelIntro}A).}
\tcb{
In the process of the simulation, these cells can acquire new pathway mutations among three possible (shown in Fig. \ref{fig:ModelIntro}A).
}
If the pathway is mutated, the component is colored red and is marked with `X'.
\sout{A single mutation is sufficient for the cell to be considered as a mutant cell.}
\tcb{
% I'm not a fan of "hypermutated", but I'll use it for now so that a quick replace all can change it later.
We distinguish between cells that have zero of the three pathway mutations and those that have at least one by referring to the latter as hypermutated.
}
\par
We model immune cells as continuous variables, \sout{appropriate since at the level of immune cells it is not necessary to keep track of single cell effects.}
\tcb{
effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
}
The cytokine TGF-$\beta$ is \sout{\tcr{continuous} within the TME, i.e. the system is well-mixed.}
\tcb{
also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
}
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
\sout{While the score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
\tcb{While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.}
% Above line replaced
In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the \sout{color} \tcb{shade of blue} of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

\subsection{Tissue Cells}\label{TissueCells}
The tissue cells at all times have associated with them two important quantities: which combination of mutations they carry and an EMT score.
The model has three pathways that have the potential to mutate over the course of a simulation.
There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
The EMT score will be exposited in Section \ref{EMT}.

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis. Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
The NKs and CTLs act on the system by recognizing \sout{mutated} \tcb{hypermutated} cells and clearing them.
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear \sout{mutated} \tcb{hypermutated} cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

\tcb{
\subsection{Inflammation} % How did we not have a section on this before?
We also include an inflammation score that follows a fixed cycling scheme between low and high inflammatory states.
This is to simulate the various changes to the TME that can happen over long periods of time (days and weeks).
For the purpose of the model, we consider low inflammation to be the default state.
When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
\tcr{See Table 2 in the supplement to see which parameters are changed by the inflammation score changing.} % Table 2 or S2? Change label in supplement?
}

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purpose of \tcb{parameter values} of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have \sout{some} \tcb{a subset} of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear \sout{mutated} \tcb{hypermutated} cells, the increased immune evasion only benefits \sout{mutated} \tcb{hypermutated} mesenchymal cells.
\par
\tcb{
In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
}
\subsection{Model Simulation}

\subsubsection{Initial conditions} 
Every model run starts with \sout{a fixed amount} \tcb{an amount} of \tcb{tissue} cells, $N_0$ \tcb{, determined by the set of parameters used for the run.
A set of warmup cycles are run to ensure arrival to this steady state.
During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.}
\sout{However, different parameter values will lead to different steady states for the population of tissue cells.}
\sout{In light of this, a warmup period of ~1000 cell cycles is used in which time no mutations are allowed to happen.}
\sout{Thus, the only immune cell present is NKs.}
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.

@@

\paragraph{Proliferation}
There are four factors that contribute to the weight of a cell to proliferate.
The first is a base proliferation rate that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
% Add: 
\tcb{This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.}
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
In total, the weight for proliferation is given by

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be \sout{mutated} \tcb{hypermutated} ($\delta_{\text{MUT}}=1$).
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
In total, the weight of NK clearance is given by

@@

\paragraph{Rest in $G_0$} 
\sout{The weight associated with quiescence is taken as 1 except in the case of mesenchymal cells.
For mesenchymal cells, the weight lost to proliferation is added to the weight of quiescence.
Hence, the weight of quiescence is given by}
\tcb{The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
Hence, the weight of rest is given by}
% Replace quiescence with rest and explain more above

@@

\tcb{Again, t}\sout{T}he reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
\par
Finally, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common. So, this is not really so much a TGFB received as it is a change in TGFB signaling.
\sout{Each cell receives an amount of TGF-$\beta$ given by}
\tcb{The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.}

@@

\tcb{
We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
Finally, if this quantity is large enough, \sout{the cell undergoes EMT; otherwise, it undergoes MET.} \tcb{the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.}
}
% Need to add better explanation for Hill fn
\sout{The summand on the left deterministically divides up the total TGF-$\beta$ among all $N$ cells evenly and a Hill function on this quantity determines how much TGF-$\beta$ each cell thus receives.}
\sout{Then, some randomness is introduced by the independent and identically distributed variables $X_i$ where $\sigma$ is a parameter in the model.
If this quantity is large enough, relative to the EMT score of the cell, then the cell will undergo EMT and its EMT score will increase.
If not, then the cell undergoes the reverse process, MET, and its EMT score will decrease.
}
Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.

@@

For CTLs and Tregs, they rely on \sout{mutant} \tcb{hypermutated} cells being cleared before they can be activated.
Let $N_{\text{MUT}}^*(k)$ represent the number of \sout{mutant} \tcb{hypermutated} cells cleared by the immune system during cell cycle $k$.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
We choose the following differential equations to govern the CTL and Treg populations:

@@

\begin{figure}
\center
{\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of agent-based model components; \tcb{
each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
}\sout{blue/red denote tissue cells with/without mutations} \tcb{red/blue denotes mutated/unmutated pathways} . Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
 %The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: \sout{mutated} \tcb{hypermutated} cells; Mes: mesenchymal cells. 
C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
\label{fig:ModelIntro}
\end{figure}

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the \tcr{probability} of cancer, and the Time to Cancer. 

@@

The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear \sout{mutated} \tcb{hypermutated} tissue cells, and deactivate upon successfully carrying out their cytotoxic function.

@@

\sout{
However, after approximately 1000 cell cycles (750 days), one mutated cell exhibits clonal growth, and a tumor is established.
At this point, large numbers of cells from the adaptive immune system (CTLs and Tregs) are being heavily recruited, and a peak in TGF-$\beta$ expression is observed.
The number of mesenchymal cells increases initially with increasing TGF-$\beta$, but decreases after the clone takes over the tissue.
This shift back towards an epithelial tissue phenotype occurs even though the expression of TGF-$\beta$ remains high.
After 841 cell cycles, the proportion of mutated cells reaches 50\%: this is modeled as the threshold defining tumorigenesis. Thus this patient has a Time to Cancer of 841 cell cycles, or 631 days. 
After cancer incidence, the number of mutated cells continues to grow rapidly and soon makes up 100\% of the cell population. A peak in mesenchymal phenotypes is observed following cancer incidence, before most transition back to an epithelial state.}
\tcb{
After about 700 cell cycles, the hypermutated cell population begins to grow steadily.

@@

After 841 cell cycles, the proportion of hypermutated cells reaches 50\%: the threshold defining progression.

@@

For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of hypermutated cells grows rapidly and soon makes up 100\% of the cell population.

@@

\tcr{THE REST OF THIS SECTION IS FROM THE FIRST DRAFT AND WILL BE DELETED}

>>>>

\section*{Abstract}
During cancer progression, the immune system is engaged in complex sets of interactions with epithelia that affect tumor prognosis.
Inflammation also has significant effects on gastrointestinal and pancreatic tumors.
Moreover, epithelial cell plasticity can significantly alter disease trajectories (for better or worse) via epithelial-to-mesenchymal transition (EMT).
Several of the same pathways that regulate EMT are involved in tumor-immune interactions, yet little is known about the mechanisms and consequences of these regulatory processes.
Here we introduce a multiscale evolutionary model to describe the interplay between tumor-immune-EMT interactions and their impact on epithelial tissues.
Through {\em in silico} analysis of large patient cohorts, we found controllable regimes that maximize cancer-free survival.
Delaying progression depended crucially on properties of the mesenchymal phenotype: its growth rate and its immune-evasiveness.
We performed analysis of pancreatic cancer data from The Cancer Genome Atlas to compare clinical measurements with model predictions.
We found, in broad agreement with the model, that association with EMT worsened the survival probabilities of inflammation-associated cancer.
These results offer novel means to delay disease progression by regulating properties of EMT, and demonstrate the importance of studying cancer-immune interactions in light of EMT.

@@

In addition, the immune system plays important regulatory roles in premalignant tissue, where inflammation can increase the probability of tumorigenesis (and/or decrease the time to cancer), including in tumors originating in gastrointestinal and pancreatic tissues \cite{hu10_inflammationinduced, balkwill01_inflammation}.

@@

Recent work raises some questions about the relationship between inflammation and cancer, suggesting that under certain conditions inflammation may not be oncogenic but rather onco-protective \cite{guo17_multiscale}.

@@

While EMT has for some time been assumed to impact metastasis, more recently EMT-related phenotypes have been linked to other aspects of both tumor initiation \cite{puisieux2014oncogenic} and progression \cite{nieto2016emt,peinado2007snail}.

@@

There is evidence directly linking Treg-secreted TGF-$\beta$ with EMT in hepatocellular carcinoma\cite{shi2019cd4+}.
Thus, even by considering only a single signaling factor, we find these three components (the tumor, the immune system, and EMT) to be linked. It therefore strikes us as a priority to develop models to understand how interactions between each of these three components affect cancer incidence and progression.

@@

While many cell behaviors are associated with stemness, here we target the specific reduction in proliferation rate \cite{woods2014effects}.

@@

We seek to describe a set of crucial molecular and cellular interactions in epithelial tissue cells, including effects due to DNA damage and mutation, to investigate the probability that premalignant tissue will progress and, if so, when.

@@

% commented to keep same line numbers as Paper.txt (as of 9/16/19)

@@

% commented to keep same line numbers as Paper.txt (as of 9/16/19)

@@

Global sensitivity analysis identifies parameters that are crucial for progression.
We then study these in more depth, focusing on the competing effects of EMT and of the immune system on progression: we find that EMT intricately regulates progression and that under certain regimes a careful balance of EMT- and immune-driven processes can significantly prolong cancer-free survival.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)

@@

\begin{table}[H]
\begin{tabular}{|p{\textwidth}|}
\hline
%%%
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
\begin{itemize}
\item TGF-$\beta$ is well-mixed within the TME
\item All cells are on the same ``clock'', i.e. they all decide at the same time what to do next
\item During a given cell cycle, exactly one event can happen to any one cell: proliferation, apoptosis, immune clearance, or rest
\item All tissue cells (mutated and non-mutated) are equally likely to encounter immune cells
\item Other spatial effects are ignored by the model, e.g. angiogenesis, hypoxia, invasive front, and chemotaxis
\item The initial cells are mutated, but are not able to form an aggressive cancer without further mutations
\end{itemize}
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
\\
\hline
\end{tabular}
\caption{\tcb{Basic assumptions of the model}}
\label{table:model_assumptions}
\end{table}

\section{Methods}
We develop an agent-based model to describe the relationships between cancer, the immune system, and EMT, building on cell-cycle and tissue-cell components of \cite{guo17_multiscale}. 
% might want to define "driver mutation" in intro. See PreMalignant_Thoughts.txt for definition found in a source
The agents in the model are the tissue cells that have already formed a premalignant tumor yet lack key driver mutations.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
In the process of the simulation, these cells can acquire driver mutations in any of three key pathways (shown in Fig. \ref{fig:ModelIntro}A).
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
If the pathway is mutated, the component is colored red and is marked with `X'.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% I'm not a fan of "malignant", but I'll use it for now so that a quick replace all can change it later.
We distinguish between cells that have zero driver mutations and those with at least one as premalignant and malignant, respectively.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
\par
We model immune cells as continuous variables, 
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
effectively assuming that the TME is well-mixed with regards to the immune infiltrate.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
The cytokine TGF-$\beta$ is
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
also assumed to be well-mixed in the TME so that all tissue cells have equal access to it.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
Tissue cells can take on either epithelial or mesenchymal phenotypes in a plastic manner: these phenotypes depend on both the TME and cell-intrinsic factors.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
While the EMT score is continuous, a threshold determines if a given cell is labeled as epithelial or mesenchymal.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
In Fig. \ref{fig:ModelIntro}A, the EMT labeling is given by the shade of blue of the cell.
Even though each cell does have an EMT score on a spectrum of values, only the labeling influences its interactions in the model.

\subsection{Tissue Cells}\label{TissueCells}
The tissue cells at all times have associated with them two important quantities: which combination of driver mutations they carry and an EMT score.
The model has three pathways that have the potential to mutate over the course of a simulation.
There is a proliferation pathway that when mutated increases the probability of the cell proliferating.
There is an apoptosis pathway that when mutated decreases the probability of the cell undergoing apoptosis.
There is an immune evasion pathway that when mutated decreases the probability that an immune cell can effectively clear the mutated cell.
The EMT score will be exposited in Section \ref{EMT}.

\subsection{The Immune System}\label{ImmuneSystem}
The immune system is modeled as having three active immune cell types: NKs, CTLs, and Tregs.
% We could justify the immune system not clearing "base" mutated cells by saying the immune surveillance provides a baseline control over these cells accounted for in their apoptosis. Or suggesting that we are interested in immune activity above and beyond the baseline requirement to keep this tumor in check.
The NKs and CTLs act on the system by recognizing malignant cells and clearing them.
Upon successful clearance, they themselves are then deactivated and removed from the immune population.
Tregs act on the system by suppressing the effector functions of NKs and CTLs, making it less likely they can clear malignant cells.
In addition to this, Tregs release TGF-$\beta$ which further shapes the TME by pushing tissues cells more towards a mesenchymal phenotype.

% commented to keep same line numbers as Paper.txt (as of 9/16/19)
\subsection{Inflammation} % How did we not have a section on this before?
We also include an inflammation score that follows a fixed cycling scheme between low and high inflammatory states.
This is to simulate the various changes to the TME that can happen over long periods of time (days and weeks).
For the purpose of the model, we consider low inflammation to be the default state.
When the inflammation switches to high, then a number of the parameters change values to reflect the additional immune activity.
See Table 2 in the supplement to see which parameters are changed by the inflammation score changing. % Table 2 or S2? Change label in supplement?
% commented to keep same line numbers as Paper.txt (as of 9/16/19)

\subsection{EMT}\label{EMT}
Each tissue cell has an EMT score between 0 and 1.
When the score is above a fixed threshold value, the cell is labeled as mesenchymal.
Otherwise, it is below the threshold and labeled as epithelial.
For the purpose of parameter values of the model, a cell in an epithelial state is considered in the base state, and one in a mesenchymal state will have a subset of its parameters updated.
In particular, mesenchymal cells benefit from increased immune evasion but suffer from decreased proliferation.
Note, that because the immune system can only clear malignant cells, the increased immune evasion only benefits malignant mesenchymal cells.
\par
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
In modeling EMT this way, we are assuming that the same factor, TGF-$\beta$, drives EMT both at initiation and through progression of cancer.
In addition, as per our model assumptions (see Table \ref{table:model_assumptions}), we choose to ignore other factors that are known to induce EMT such as hypoxia as well as the typical spatial organization of epithelial and mesenchymal cells as this pertains to the invasive front.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
\subsection{Model Simulation}

\subsubsection{Initial conditions} 
Every model run starts with an amount of premalignant tissue cells, $N_0$, determined by the set of parameters used for the run.
A set of warmup cycles are run to ensure arrival to this steady state.
During these warmup cycles, no mutations are allowed to occur the only immune cells present are NKs.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
After the warmup cycles are complete, a mutagenic event is simulated in which all the tissue cells experience a random increase in their probability to mutate.
From then on, a cell which proliferates also either mutates or increases its probability of mutating later on.

@@

\paragraph{Proliferation}
There are four factors that contribute to the weight of a cell to proliferate.
The first is a base proliferation rate that all cells have, $p$.
Second, if the cell has a mutation in the proliferation pathway ($\delta_P=1$), then the weight for proliferation is proportionally increased by $\Delta_P$.
Third, if the cell is mesenchymal ($\zeta=1$), then the weight for proliferation is proportionally decreased by $\Delta_{\text{MGA}}$, which stands for mesenchymal growth arrest.
% Add: 
This lost proliferation for mesenchymal cells will later be used to increase their chance of resting.
Fourth, there is a negative feedback of the cells on their own proliferation which is quantified by a Hill factor as a function of the tissue cell population, $N_C$, with EC50 term $K_0$.
In total, the weight for proliferation is given by

@@

\paragraph{Immune Clearance}
For both NK clearance and CTL clearance, the weights are built with the same factors but have different parameter values for NK and CTLs.
First of all, the cell needs to be malignant ($\delta_{\text{MUT}}=1$).
Second, there is a Hill factor that captures the probability of an immune cell finding and interacting with the given tissue cell with EC50 term $K_1$.
Third, NKs and CTLs have their own efficacy parameters, $E_{\text{NK}}$ and $E_{\text{CTL}}$, which can be understood as the rate of immune clearance given an immune cell has found the mutated cell.
Fourth, there is a decreasing Hill factor based on the number of Treg cells present with EC50 term $K_2$.
Finally, there are two factors that proportionally decrease the weight of immune clearance depending on if the cell has an immune evasion mutation ($\delta_{\text{IE}}=1$) or if it is mesenchymal ($\zeta=1$) with respective decreases $\Delta_{\text{IE}}$ and $\Delta_{\text{MIE}}$.
In total, the weight of NK clearance is given by

@@

\paragraph{Rest in $G_0$} 
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
The weight associated with rest is taken as 1 except in the case of mesenchymal cells.
Recall that mesenchymal cells had their proliferation rate decreased by $1-\zeta\Delta_{\text{MGA}}$ (see Eq. 2.1).
The biological assumption here is that mesenchymal cells instead of proliferating will instead rest, so this lost proliferation weight is added to the resting weight.
Hence, the weight of rest is given by
% Replace quiescence with rest and explain more above

@@

Again, the reason for adding that term is due to the understanding that overall mesenchymal cells proliferate less as individual cells rest longer in the $G_0$ phase.

\subsubsection{Completing the Cell Cycle}
After the cell fates are determined and the results reflected in the system, there are a few things that happen before the system moves on to a new cell cycle.
First, the NK and CTL populations are reduced by the number of mutated cells they cleared.
This represents the fact that individual immune cells lose efficacy as they carry out their effector functions. 
Second, all proliferating cells have a cell-specific probability of undergoing a pathway mutation.
If they do, one is randomly chosen among the three pathways and the pathway in that cell mutates.
If the cell does not undergo a mutation, then its probability of mutation during subsequent cell cycles increases.
\par
Finally, the EMT values for each cell is updated.
This depends on the cells current EMT score and how much TGF-$\beta$ is currently in the system.
% The following sentence could be altered to reflect that the threshold value for $\tau$ is 0, so not only are negative values possible, but very common. So, this is not really so much a TGFB received as it is a change in TGFB signaling.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
The amount of TGF-$\beta$ absorbed by all cells is given by an increasing Hill function in terms of the TGF-$\beta$ in the TME.
The saturation effect is to limit the amount of TGF-$\beta$ a cell can absorb in a given time interval.
This quantity is then divided up randomly among the $N_C$ living cells via a normally distributed noise term to determine how much exogenous TGF-$\beta$ each cell receives during this cell cycle.
Should this value, $\tau_i$ in Eq. 2.5, be negative, we interpret this as the cell losing TGF-$\beta$ to the TME and thus being more likely to undergo MET.

@@

% commented to keep same line numbers as Paper.txt (as of 9/16/19)
We then combine $\tau_i$ with the current EMT score of the cell, as a proxy for the endogenous TGF-$\beta$.
Finally, if this quantity is large enough, the EMT score of the cell increases towards 1; otherwise, it decreases towards 0.
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% Need to add better explanation for Hill fn
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
Each cell then is relabeled as either epithelial or mesenchymal depending on its new EMT score and whether it is below or above the mesenchymal threshold.
Thus, there are two main factors that determine if a cell will end a cell cycle as mesenchymal: concentration of TGF-$\beta$ in the system and the current EMT score of the cell.

@@

For CTLs and Tregs, they rely on malignant cells being cleared before they can be activated.
Let $N_{\text{MUT}}^*(k)$ represent the number of malignant cells cleared by the immune system during cell cycle $k$.
In addition, Treg recruitment is upregulated by TGF-$\beta$, which will be incorporated via a Hill function with EC50 term $K_4$.
We choose the following differential equations to govern the CTL and Treg populations:

@@

\begin{figure}
\center
{\includegraphics[width=0.9\textwidth]{Figure1/Figure1.pdf}}
\caption{A. Schematic depiction of agent-based model components;
each of the 10 columns represents a single tissue cell divided into three compartments representing the state (mutated or not) of the three pathways with mutagenic potential;
red/blue denotes mutated/unmutated pathways. Black arrows depict regulation of the cell fate in each cell cycle. Inset depicts major interactions between the immune system and tissue cells.
 %The parameters used in this model can be found in the Supplementary material. Unless otherwise stated, parameter values are assumed to be those found in the Supplement. \tcr{Thoughts on the wording here? Once I have the table, I will use the name of that table.}
B. A representative simulation of one patient. The model parameter values used can be found in Supplementary Table 2. 
The inflammation cycling scheme is represented above the patient dynamics. The vertical dashed line denotes the end of the warmup period. Mut: malignant cells; Mes: mesenchymal cells. 
C. Survival curve for one cohort of patients with the parameter values given in Supplementary Table 2. }
\label{fig:ModelIntro}
\end{figure}

We begin by investigating general features of the model to establish baseline conditions and to assess the impact of various model components on the key measured outcomes: the probability of progression, and the Time to Cancer. 

@@

The different means by which the immune system acts on tissue cells are also shown (\ref{fig:ModelIntro}A Inset). NK cells and CTLs attempt to clear malignant tissue cells, and deactivate upon successfully carrying out their cytotoxic function.

@@

% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
% commented to keep same line numbers as Paper.txt (as of 9/16/19)
After about 700 cell cycles, the malignant cell population begins to grow steadily.

@@

After 841 cell cycles, the proportion of malignant cells reaches 50\%: the threshold defining progression.

@@

For this patient, we kept let the simulation go beyond this timepoint, and we se that the number of malignant cells grows rapidly and soon makes up 100\% of the cell population.

@@
